Type-Safe Modular Parsing

### 2016-11-16 paper/paper.tex
Modification
Some text and half of the table
Text
Organize.
More.
Done text for methodology and library.

\lstset{ %
language=JavaScala,                % choose the language of the code
columns=flexible,
lineskip=-1pt,
basicstyle=\ttfamily\small,       % the size of the fonts that are used for the code
numbers=none,                   % where to put the line-numbers
numberstyle=\ttfamily\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it's 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
morekeywords={var},
%  frame=single,                   % adds a frame around the code
tabsize=2,                  % sets default tabsize to 2 spaces
captionpos=none,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
escapeinside={(**}{**)},          % if you want to add a comment within your code
keywordstyle=\ttfamily\bfseries,
% commentstyle=\color{Gray},
% stringstyle=\color{Green}
}

@@

%{\theorembodyfont{\sffamily} \newtheorem{theorem}{Theorem}}
%{\theorembodyfont{\sffamily} \newtheorem{lemma}{Lemma}}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newenvironment{proof}{\textbf{Proof:\hspace{4mm}}}{$\Box$}
\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\yanlin[1]{\authornote{yanlin}{purple}{#1}}
%\newcommand{\hl}[1]{\textcolor{red}{#1}}
\newcommand\marco[1]{\authornote{marco}{blue}{#1}}
\newcommand\haoyuan[1]{\authornote{haoyuan}{cyan}{#1}}

@@

\input{sections/Library/OpenRecursion.tex}
\input{sections/Library/ObjectAlgebras.tex}
\input{sections/Library/ParsingLibrary.tex}
\input{sections/Library/FirstTry.tex} % mention again | and ||| with concrete examples. (pApp | pVar)
\input{sections/Library/DifferentSyntax.tex}

% Object Algebras
% type synonyms, Parser[Expr], Fix[..]
% more: two pretty-printers OldPretty and NewPretty, they are separated to show modularity.
% subsection: modify the code using List[E] and type bounds.
% more subsections: lexing, traits, operations, language composition (more extensibility)

%...

>>>>

\lstdefinelanguage{PlainCode}{}

\lstset{ %
language=JavaScala,                % choose the language of the code
columns=flexible,
lineskip=-1pt,
basicstyle=\ttfamily\small,       % the size of the fonts that are used for the code
numbers=none,                   % where to put the line-numbers
numberstyle=\ttfamily\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it's 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
morekeywords={var, lazy},
%  frame=single,                   % adds a frame around the code
tabsize=2,                  % sets default tabsize to 2 spaces
captionpos=none,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
escapeinside={(**}{**)},          % if you want to add a comment within your code
keywordstyle=\ttfamily\bfseries,
% commentstyle=\color{Gray},
% stringstyle=\color{Green}
}

@@

%{\theorembodyfont{\sffamily} \newtheorem{theorem}{Theorem}}
%{\theorembodyfont{\sffamily} \newtheorem{lemma}{Lemma}}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newenvironment{proof}{\textbf{Proof:\hspace{4mm}}}{$\Box$}
\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\yanlin[1]{\authornote{yanlin}{purple}{#1}}
%\newcommand{\hl}[1]{\textcolor{red}{#1}}
\newcommand\huang[1]{\authornote{huang}{blue}{#1}}
\newcommand\haoyuan[1]{\authornote{haoyuan}{cyan}{#1}}

@@

\input{sections/PackratParsers.tex}

\input{sections/OpenRecursion.tex}

\input{sections/ObjectAlgebras.tex}
\input{sections/ParsingLibrary.tex}

% Object Algebras
% type synonyms, Parser[Expr], Fix[..]
% more: two pretty-printers OldPretty and NewPretty, they are separated to show modularity.
% subsection: modify the code using List[E] and type bounds.
% more subsections: lexing, traits, operations, language composition (more extensibility)
% use paragraphs
%...

### 2016-11-16 paper/sections/ObjectAlgebras.tex
Done text for methodology and library.

build(new LitPrint(){})

@@

\begin{lstlisting}
trait ExprLitParser[E] {
    val pE : (ExprAlg[E] with LitAlg[E]) => Fix[Parser[E]] = alg => p => {
        val pExprE = new ExprParser[E](){}.pE(alg)(p)
        val pLitE = new LitParser[E](){}.pE(alg)(p)
        pExprE ||| pLitE
    }
}
\end{lstlisting}

In client code, a user can either define a new function with type \lstinline{(ExprAlg[E] with LitAlg[E]) =>} \lstinline{Parser[E]} based on \lstinline{pE}, which implies an algebra is applied after \lstinline{fix}, or simply feed an algebra to \lstinline{pE} at first, then obtain the result from \lstinline{fix}:
\begin{lstlisting}
trait ExprLitPrint extends ExprPrint with LitPrint
val parsePrint = fix(new ExprLitParser[String](){}.pE(new ExprLitPrint(){}))
val result = runParser(parsePrint)("x 3") // prints "(x 3)"
\end{lstlisting}
\haoyuan{Is there a way to avoid defining ExprLitPrint?}

@@

trait LamParser[E, T] {
    type Pair = (Parser[E], Parser[T])
    val pE : LamAlg[E, T] => (=> Pair) => Parser[E] = alg => p =>
        ("\\" ~> ident) ~ (":" ~> p._2) ~ ("." ~> p._1)
            ^^ { case x ~ t ~ e => alg.lamE(x, t, e) }
    val pT : LamAlg[E, T] => (=> Pair) => Parser[T] = alg => p =>
        "Int" ^^ { _ => alg.intT() }
}
\end{lstlisting}
\lstinline{LamParser} contains parsers for both expressions and types. From the code we observe that the multi-sorts of Object Algebra interface
exactly express multiple syntax. At this point, to combine these two small components together we need some ``glue code''. But instead of writing composition by hand, we reimplement the combinator \lstinline{|||}, to compose one parser with type \lstinline{Alg1 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} and another one with type \lstinline{Alg2 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} using alternative. Note that we can only abstract over the whole algebra interface type, instead of working on its type parameters, as the number of type parameters is not fixed. In summary, \lstinline{"|||"} produces a combined parser, which takes a compound algebra as its parameter:
\begin{lstlisting}
implicit class Combinator[Alg1, Alg2, R, E](x : Alg1 => (=> R) => Parser[E]) {
    def |||(y : Alg2 => (=> R) => Parser[E]): (Alg1 with Alg2) => (=> R) => Parser[E]
        = alg => l => x(alg)(l) ||| y(alg)(l)
}
\end{lstlisting}
Now it is sufficient to merge \lstinline{ExprParser} and \lstinline{LamParser} together:
\begin{lstlisting}
trait ExprLamParser[E, T] {
    val pExpr = new ExprParser[E](){}
    val pLam = new LamParser[E, T](){}
    val pExprE : ExprAlg[E] => (=> pLam.Pair) => Parser[E] =
        alg => p => pExpr.pE(alg)(p._1)

>>>>

build(new LitPrint{})

@@

\begin{lstlisting}
trait ExprLitParser[E] {
    val pE : (ExprAlg[E] with LitAlg[E]) => Fix[Parser[E]] = alg => p => {
        val pExprE = new ExprParser[E]{}.pE(alg)(p)
        val pLitE = new LitParser[E]{}.pE(alg)(p)
        pExprE ||| pLitE
    }
}
\end{lstlisting}

In client code, a user can either define a new function with type \lstinline{(ExprAlg[E] with LitAlg[E])} \lstinline{=>} \lstinline{Parser[E]} based on \lstinline{pE}, which implies an algebra is applied after \lstinline{fix}, or simply feed an algebra to \lstinline{pE} at first, then obtain the result from \lstinline{fix}:
\begin{lstlisting}
val parsePrint = fix(new ExprLitParser[String]{}.pE(new ExprPrint with LitPrint{}))
val result = runParser(parsePrint)("x 3") // prints "(x 3)"
\end{lstlisting}

@@

trait LamParser[E, T] {
    type Pair = (Parser[E], Parser[T])
    val pE : LamAlg[E, T] => (=> Pair) => Parser[E] = alg => p =>
        ("\\" ~> ident) ~ (":" ~> p._2) ~ ("." ~> p._1)
            ^^ { case x ~ t ~ e => alg.lamE(x, t, e) }
    val pT : LamAlg[E, T] => (=> Pair) => Parser[T] = alg => p =>
        "Int" ^^^ alg.intT
}
\end{lstlisting}
Note that we are omitting its lexer here. \lstinline{LamParser} contains parsers for both expressions and types. From the code we observe that the multi-sorts of Object Algebra interface
exactly express multiple syntax. At this point, to combine these two small components together we need some ``glue code''. But instead of writing composition by hand, we reimplement the combinator \lstinline{|||}, to compose one parser with type \lstinline{Alg1 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} and another one with type \lstinline{Alg2 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} using alternative. Note that we can only abstract over the whole algebra interface type, instead of working on its type parameters, as the number of type parameters is not fixed. In summary, \lstinline{"|||"} produces a combined parser, which takes a compound algebra as its parameter:
\begin{lstlisting}
implicit class Combinator[Alg1, Alg2, R, E](x : Alg1 => (=> R) => Parser[E]) {
    def |||(y : Alg2 => (=> R) => Parser[E]): (Alg1 with Alg2) => (=> R) => Parser[E]
        = alg => l => x(alg)(l) ||| y(alg)(l)
}
\end{lstlisting}
Now it is sufficient to merge \lstinline{ExprParser} and \lstinline{LamParser} together:
\begin{lstlisting}
trait ExprLamParser[E, T] {
    val pExpr = new ExprParser[E]{}
    val pLam = new LamParser[E, T]{}
    val pExprE : ExprAlg[E] => (=> pLam.Pair) => Parser[E] =
        alg => p => pExpr.pE(alg)(p._1)

### 2016-11-16 paper/sections/ParsingLibrary.tex
Done text for methodology and library.

@@

@@

>>>>

The library is made up of around twenty lines of code only, as shown in Figure~\ref{fig:sourcelibrary}. It consists of some types wrapped up by synonyms, basic functions \lstinline{fix} and \lstinline{runParser}, and the combinator \lstinline{|||} from Section~\ref{subsec:differentsyntax}.

\begin{figure}[htbp]
\centering
\begin{lstlisting}
import scala.util.parsing.combinator._
import scala.util.parsing.combinator.syntactical._

object Library extends StandardTokenParsers with PackratParsers {
    type Fix[T] = (=> T) => T
    type Parser[T] = PackratParser[T]
    type OpenParser[Alg, List, T] = Alg => (=> List) => Parser[T]

    def fix[A](f: Fix[A]): A = { lazy val a: A = f(a); a }
    def runParser(p : Parser[_]) : String => Unit = in => {
        phrase(p)(new lexical.Scanner(in)) match {
            case t if t.successful => println(t.get)
            case t                 => scala.sys.error(t.toString)
        }
    }

    implicit class Combinator[Alg1, Alg2, R, E](x : Alg1 => (=> R) => Parser[E]) {
        def |||(y : Alg2 => (=> R) => Parser[E])
            : Alg1 with Alg2 => (=> R) => Parser[E]
                = alg => l => x(alg)(l) ||| y(alg)(l)
    }
}
\end{lstlisting}
\caption{Source code of our library.}\label{fig:sourcelibrary}
\end{figure}

@@

Having pointed out that our goal is to realize modular and extensible parsers, we recommend users to write code in a modular and
organized way. The framework we use for a language, or rather a language component in our case study, consists of the following three
parts:
\begin{itemize}
\item \textbf{Object Algebra interface:} defined as a trait with several cases, corresponding to the grammar rules.
\item \textbf{Parser:} implementation of the parser, using open recursion and Object Algebras, with its scoping trait abstracted over the type parameters.
\item \textbf{Algebra:} some behaviors (operations) on the data structure, such as pretty-printing and so on.
\end{itemize}

Figure~\ref{fig:objectexpr} shows how to encapsulate \lstinline{ExprAlg} and its related stuff in object \lstinline{Expr} using our library. Note that we use type bounds to address the issue remained in Section~\ref{subsec:differentsyntax}. \lstinline{List[E]} is a type synonym for a record type, while the type bound \lstinline{L} \lstinline{<:} \lstinline{List[E]} in \lstinline{AlgParser} states that the fix-point of our parser is a record, which contains at least \lstinline{pE} for parsing expressions. Hence it saves us from updating the type of fix-points all the time.

\begin{figure}[htbp]
\centering
\begin{lstlisting}
object Expr {
    trait Alg[E] {
        def varE(x : String) : E
        def appE(e1 : E, e2 : E) : E
    }

    type List[E] = { val pE : Parser[E] }

    trait AlgParser[E, L <: List[E]] {
        val pExprE : OpenParser[Alg[E], L, E] = alg => p => {
            lazy val pE = p.pE
            ident ^^ alg.varE |||
            pE ~ pE ^^ { case e1 ~ e2 => alg.appE(e1, e2) }
        }
    }

    trait Print extends Alg[String] {
        def varE(x : String) = x
        def appE(e1 : String, e2 : String) = "(" + e1 + " " + e2 + ")"
    }
}
\end{lstlisting}
\caption{Object \lstinline{Expr} for expressions.}\label{fig:objectexpr}
\end{figure}

Now \lstinline{Expr} turns out to be a self-contained small language with parsing and pretty-printing. To construct a different language component, we implement \lstinline{LamAlg} in Figure~\ref{fig:objectlam}, where we declare a different \lstinline{List} for multiple syntax.

\begin{figure}[htbp]
\centering
\begin{lstlisting}
object Lam {
    trait Alg[E, T] {
        def intT() : T
        def lamE(x : String, t : T, e : E) : E
    }

    type List[E, T] = { val pE : Parser[E]; val pT : Parser[T] }

    trait AlgParser[E, T, L <: List[E, T]] {
        lexical.reserved += ("Int")
        lexical.delimiters += ("\\", ":", ".")
        val pLamE : OpenParser[Alg[E, T], L, E] = alg => p =>
            ("\\" ~> ident) ~ (":" ~> p.pT) ~ ("." ~> p.pE) ^^
                { case x ~ t ~ e => alg.lamE(x, t, e)}
        val pLamT : OpenParser[Alg[E, T], L, T] = alg => p =>
            "Int" ^^^ alg.intT
    }

    trait Print extends Alg[String, String] {
        def intT() = "Int"
        def lamE(x : String, t : String, e : String) = "\\" + x + ":" + t + "." + e
    }
}
\end{lstlisting}
\caption{Object \lstinline{Lam} for lambdas.}\label{fig:objectlam}
\end{figure}

@@

We have optimized the framework for not only modular parsers, but also languages, and that is the reason why we call them language components. To achieve
higher-order modularity, it is necessary to follow the structure when composing two small components. Figure~\ref{fig:objectlamexpr} presents the combination of \lstinline{Expr} and \lstinline{Lam}, as a new language, which is again modular for future compositions. Note that the \lstinline{AlgParser} extends both parent parsers, so as to reuse their parser functions, as well as composing lexers automatically.

\begin{figure}[htbp]
\centering
\begin{lstlisting}
object LamExpr {
    type List[E, T] = Lam.List[E, T]
    type Alg[E, T] = Expr.Alg[E] with Lam.Alg[E, T]
    trait Print extends Expr.Print with Lam.Print

    trait AlgParser[E, T, L <: List[E, T]] extends Expr.AlgParser[E, L]
            with Lam.AlgParser[E, T, L] {
        val pLamExprE = pExprE ||| pLamE
        val pLamExprT = pLamT
    }
}
\end{lstlisting}
\caption{Object \lstinline{LamExpr}, composing \lstinline{Lam} and \lstinline{Expr} together.}\label{fig:objectlamexpr}
\end{figure}

Below we give an example of client code for testing \lstinline{LamExpr}. \lstinline{parse} is a generic parser generator, which takes an algebra.
In \lstinline{parseAndPrint} we pass the printing algebra to \lstinline{parse}, so that it parses the input and apply pretty-printing, returning the result as expected.

\begin{lstlisting}
class List[E, T](pe : Parser[E], pt : Parser[T]) {
    val pE = pe; val pT = pt
}
def parse[E, T](inp: String)(alg: LamExpr.Alg[E, T]) = {
    def parser : Fix[List[E, T]] = l => {
        val lang = new LamExpr.AlgParser[E, T, List[E, T]] {}
        new List[E, T](lang.pLamExprE(alg)(l), lang.pLamExprT(alg)(l))
    }
    runParser(fix(parser).pE)(inp)
}
def parseAndPrint(inp: String) = parse(inp)(new LamExpr.Print {})

val result = parseAndPrint("\\x:Int.x y") // prints "\\x:Int.(x y)"
\end{lstlisting}
